Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-05-24 analysis/paper/article.qmd
resubmission complete

Such challenges to the autonomy and self-governance in higher education are particularly remarkable given that many architects of the current administration hold degrees from the very institutions they are attempting to dismantle, from Donald Trump (Wharton School of Business), to J.D. Vance (Yale Law School), Chris Rufo (Georgetown), Stephen Miller (Duke University), and Peter Hegseth (Harvard Kennedy School). Reviewing the profiles of the board of trustees for the Heritage Foundation—the conservative think tank which generated the Project 2025 report that has been used as a blueprint for feeding U.S. higher education "into a woodchipper" (Kunder 2025)—is also illustrative. Two thirds of the trustees on the Heritage Foundation board have advanced degrees, while one third have doctoral degrees from institutions such as Harvard, Oxford, the University of Colorado, and the University of Texas. Lindsey Burke, who authored the Project 2025 chapter on the Department of Education, holds a PhD from George Mason University. Such career trajectories are a testament to the elite chiasmus that decries the merits and value of higher education while leveraging the credentials granted by elite institutions for professional gain. As Ho notes of the parallel systems of credentialing that undergird the ideology of "smartness" among Wall Street investment banks, playing the role of "master of the universe" requires both "especially strong doses of self-confidence and institutional legitimation" [-@ho2009, pp. 41]. Even archaeology is susceptible to the power and pull of such prestige hierarchies, which influence everything from hiring networks [@speakman2018; @kawaSocialNetworkUS2019] to publishing decisions [@beck2021] to research design[@wobst1983].

@@

Second, our results show that hiring in academic archaeology is responsive to larger cultural and political shifts. The increased popularity of Indigenous and historical archaeology from 2019–2021 reflects the prominence and impact of movements such as Black Lives Matter and a then-growing interest in conversations about race and national history. The continual responsiveness to current events revealed by our analysis of archaeology job ads demonstrates that our narratives of the past are entangled in the social and political milieu in which we work. This finding is consistent with previous work—@trigger1984 outlined how archaeological research programs are shaped by national histories, while @soffer1983, @blakey1983, @meltzer1983, and @wilk1985ancient have interrogated the entanglement of identity, ideology, national history, politics, and archaeology across space and time. Our study of job ads shows an active effort by archaeologists to take control of interpreting the past, offering an example of how disciplinary choices can shape archaeological research programs. That said, in early 2025 many researchers had their work disrupted by cancellations of funding from the U.S. National Science Foundation if their projects contained certain keywords [@Mervis], and others were told to cease activities if their projects contain certain words, including "cultural relevance," "institutional," "historically," "socioeconomic," and "systemic" [@johnson2025]. This unprecedented keyword vetting suggests that future topical shifts in archaeological research may be as much an individual response to top-down funding pressures [@wobst1983] as a disciplinary response to the current historical moment.

>>>>

Such challenges to the autonomy and self-governance in higher education are particularly remarkable given that many architects of the current administration hold degrees from the very institutions they are attempting to dismantle, from Donald Trump (Wharton School of Business), to J.D. Vance (Yale Law School), Chris Rufo (Georgetown), Stephen Miller (Duke University), and Peter Hegseth (Harvard Kennedy School). Reviewing the profiles of the board of trustees for the Heritage Foundation—the conservative think tank which generated the Project 2025 report that has been used as a blueprint for feeding U.S. higher education "into a woodchipper" (Kunder 2025)—is also illustrative. Two thirds of the trustees on the Heritage Foundation board have advanced degrees, while one third have doctoral degrees from institutions such as Harvard, Oxford, the University of Colorado, and the University of Texas. Lindsey Burke, who authored the Project 2025 chapter on the Department of Education, holds a PhD from George Mason University. Such career trajectories are a testament to the elite chiasmus that decries the merits and value of higher education while leveraging the credentials granted by elite institutions for professional gain. As Ho notes of the parallel systems of credentialing that undergird the ideology of "smartness" among Wall Street investment banks, playing the role of "master of the universe" requires both "especially strong doses of self-confidence and institutional legitimation" [-@ho2009, pp. 41]. Even archaeology is susceptible to the power and pull of such prestige hierarchies, which influence everything from hiring networks [@speakman2018; @kawaSocialNetworkUS2019] to publishing decisions [@beck2021] to research design [@wobst1983].

@@

Second, our results show that hiring in academic archaeology is responsive to larger cultural and political shifts. The increased popularity of Indigenous and historical archaeology from 2019–2021 reflects the prominence and impact of movements such as Black Lives Matter and a then-growing interest in conversations about race and national history. The continual responsiveness to current events revealed by our analysis of archaeology job ads demonstrates that our narratives of the past are entangled in the social and political milieu in which we work. This finding is consistent with previous work—@trigger1984 outlined how archaeological research programs are shaped by national histories, while @soffer1983, @blakey1983, @meltzer1983, and @wilk1985ancient have interrogated the entanglement of identity, ideology, national history, politics, and archaeology across space and time. Our study of job ads shows an active effort by archaeologists to take control of interpreting the past, offering an example of how disciplinary choices can shape archaeological research programs. That said, in early 2025 many researchers had their work disrupted by cancellations of funding from the U.S. National Science Foundation if their projects contained certain keywords [@Mervis], and others were told to cease activities if their projects contain keywords such as "cultural relevance," "institutional," "historically," "socioeconomic," and "systemic" [@johnson2025]. This unprecedented keyword vetting suggests that future topical shifts in archaeological research may be as much an individual response to top-down funding pressures [cf. @wobst1983] as a disciplinary response to the current historical moment.
